Do educational institutions explain SES gaps?: an interntational comparison of the SES achievement gap from 2000-2015

### 2017-10-09 paper/paper.Rnw
Updated hypothesis section + some text

Developmental psychology has been studying the subject for a long time and they find that the early stages in a child's life course are extremely important, if not the most, for cognitive development and defining personality traits \citep{duyme1999, waldfogel2006}. Economists and educators have long neglected these findings until the work of James Heckman brought our attention to it. For example, he has showed that cognitive and non-cognitive inequalities are present before entering school \citep{heckman2006}. This means that inequality starts strong before the schooling of a child begins and schools have the potential to equalize the learning ground. These inequalities can be narrowed significantly with high quality learning environments, specially for children coming from unfavorable conditions. But one of the most important findings from \citet{cunha2006} is that the cognitive level of a child in time \(t\) is a direct function of the experiences of time \(t-1\). Although this sounds logical, it's importance is not understood entirely.

The implications of this cumulative model suggests that it is much less cost-effective to invest in children in time \(t\) that in \(t-1\). That is why the whole debate about narrowing inequality between high and low SES families should start in early education -- it is cheaper than investing at later stages and it brings about the largest returns. Despite all of these findings and strategies to reduce the gaps, we still find that the relationship between children's parental education and future destination is present in virtually all empirical studies of social mobility and inequalities \citep{breen1997, breen2007, waldfogel2006, bradbury2015, chetty2016}.

For this reason, it is incomplete to study educational inequalities without considering the prevalence of intergenerational transmission of human and social capital. Social mobility has deepen our knowledge on the relationship between intergenerational transmission of SES and inequality of opportunity. The field of economics of education has also helped to understand that early high-quality education is by far the best investment that yields the highest returns. This is not only for the child and his/her family, but to society as a whole as it boosts economic activity, it helps the labor market improve job conditions and maintain a rapid economic growth \citep{hanushek2007}. Since \citet{coleman1966} we've spent most of our time studying the mechanisms through which this inequality comes about. Naturally, we want to do that in order to reverse it and help every child reach its fullest potential. Despite our good intentions, we've concentrated little on the magnitude of the gaps, specially in terms of cognitive abilities. We have very little comparative evidence on which countries have big or small cognitive gaps resulting from SES origins. And even more importantly, we haven't assessed whether policy efforts to reduce inequality have actually had an impact in reducing the achievement gap over time.

@@

From Denmark, for example, the literature has learned a great deal about how to improve social mobility. But that was possible by first learning that Denmark is by far one of the most mobile countries; this is the case specially for families of low-ses origin \citep{bjorklund2009, jaeger2007}. The important finding came when we discovered the main reason behind their social escalator: it's educational system. For example, research by \citet{esping_waldfogel2012} and \citet{bauchmuller2014}, show that the Danish early education system has important and longstanding impact on improving opportunities. The education system is completely subsidized for all children, otherwise giving opportunities to families who would not have been able to pay. In addition to this, Denmark is recognized as a world-leader in terms of public support for its child-care system as it spends around 2\% of GDP, and has among the highest enrollment rates for children under 6 years old \citep{esping_waldfogel2012}. Moreover, it separates students into different tracks very late compared to other European countries (age 16), something which has been linked to less educational inequality \citep{hanushek_woesmann_tracking}. These two traits make the Danish system uniquely effective. The early schooling experience attempts to level children to the same degree and this process is not stopped by different curricular tracks because it starts at around age 16 when cognitive abilities are less malleable. In short, the importance of this finding, is that we should step back and first study whether the effect is there, how big is it, and then proceed to find the causes behind it.

@@

However likely, this is a highly contested topic. A recent \href{http://www.epi.org/files/pdf/101972.pdf}{report} by the Economic Policy Institute finds that the black-white income gap has been growing since 1979, very similar to the increase of income inequality. However, if the achievement gap would be a function of income inequality, we should expect for the white-black achievement gap to be widening rather than narrowing, and those are not the findings given by \citet{reardon2011} and \citet{magnuson2008} \footnote{The black-white achievement gap narrowed significantly for cohorts between 1950 and 1970 where income inequality was increasing.}.

@@

\subsection{Research questions}

This study seeks to study the evolution of the high/low SES achievement gap in the past 15 years for all PISA participant countries and search for a likely explanation. This is different from previous work because it concentrates solely on 15 year old children, something not done for the achievement gap, and it attempts to capture the evolution of the achievement gap for countries with different educational systems. In addition, it extends the previous findings by studying the relationship between the gap and the average performance of each country in order to detect if there's an over-time tradeoff between good performance and equality over time separately for each country. The advantages of this study are threefold.

First, we concentrate on the evolution of the gap for only 15 year olds, which will serve as a comparison to the single year-country snapshot of \citet{anna2016} and the evolution of the kindergarten gap in \citet{reardon_portilla}. As we've seen before, there are reasons to think that specific age-groups have seen changes in the achievement gap. Moreover, in almost all countries with a tracked curriculum children are either at or in the process of tracking by the age 15, meaning that we will be able to link whether tracked countries are the most variable in their evolution of achievement gaps.

Second, we will be able, for the first time, to study the evolution of the achievement gap for several countries other than the United States. This will help assess the magnitude of the inequality as well as the rate of growth/decline. This study not only documents the size and changes of the SES gap, but pays particular attention to the source of the changes. That is, we study whether bottom performers are falling behind, if top performers are gaining advantages, or if both phenomenas are simultaneously at play.

Third, we attempt to model why the achievement gap is changing over time. In contrast to other studies, this is a novelty given that most research has been done cross-sectionally between countries.

More formally, we're interested in studying
a) Which countries are experiencing changes in the achievement gap;
b) establishing whether the gaps are widening/narrowing because particular groups are getting ahead/behind;
c) The size of the achievement in a comparative perspective and its relationship to overall achievement levels;
d) What explains this cross-country evolution.

We develop each question separately for more detail.

a) The seminal work of \citet{reardon2011} suggests that achievement gaps change, and they do so much quicker that we though. After recording a SES gap increase of about 40\% in only 30 years, \citet{reardon_portilla} stress that they also found a significant decrease in only 15 years of data, showing how important it is to study the changes in the achievement gap over time. We will compare the percentage change at which the gap widened/narrowed from the first to the last year available. This will give us a general idea of the overall change over time, and will allow us to compare our estimates to the actual literature \footnote{Although no study has performed this age-specific achievement gap for comparable tests over such a long time. Our results will serve as comparison for other studies that use age-specific groups, such as 4th graders using PIRLS.}

b) The widening/narrowing of the achievement gap has a source, which has been often studied to be related to everything from educational spending, income inequality, time allocation to students and preschool enrollment. The literature has concentrated very narrowly on whether the gap is increasing because the top performers are getting ahead, because both are distancing or because the bottom is falling behind. We shall pay particular attention to identifying the rate at which the top/bottom groups are evolving over time.

c) We want to investigate whether better performing countries have lower levels of inequality than other countries. \citet{werfhorst_mijs} emphasize that there is empirical evidence that suggests this. This pattern is not so obvious, however. For example, countries with high levels of tracking could maximize student performance, specially the high SES students, raising their overall performance and thus raising the national performance score. But if the bottom performers are not gaining at the same rate, then the achievement gap will inevitable grow resulting in a high performing countries with a widening achievement gap.

d) Tightly linked to the previous point, we want to test whether several dimensions of the tracking setup explain why the gaps are changing. This will support the previous points in suggesting that the dynamics of the over-time achievement gap are endogenously intertwined with the effect that tracking has on the specific changes of the high and low SES groups.

\subsection{Hypothesis}

Following the previous questions and the literature on educational inequality and tracking we pose three hypothesis for answering questions (b), (c) and (d) given that question (a) is purely an exploratory exercise.

\begin{itemize}
\item Hypothesis 1: In countries where there is a high degree of tracking we should expect the top and bottom to be evolving at a similar rate given that tracking is thought to maximize the learning experience of both groups. For countries with low levels of tracking we should expect the top and bottom groups to be closer to each other and for the gap to be closing since the both groups will be in the same track (this does not mean that the top is lowering their score but that the bottom is most likely catching up).

\item Hypothesis 2: Building on \citet{werfhorst_mijs}, then there should be a negative relationship between the average country performance and their level of inequality. Countries with high performance should have less inequality and countries with low performance should have greater inequality

\item Hypothesis 3: if any (or both) of the previous hypothesis are confirmed, then tracking should play an important role on the evolution of the achievement gap. We hypothesize that the degree of tracking is tightly related to changes in the gap, and the more tracking, the more inequality. Moreover, the more vocational tracking, the less inequality considering that it gives short term returns in terms of labor market opportunities.
\end{itemize}

>>>>

Developmental psychology has been studying the subject for a long time and they find that the early stages in a child's life course are extremely important, if not the most, for cognitive development and defining personality traits \citep{duyme1999, waldfogel2006}. Economists and educators have long neglected these findings until the work of James Heckman brought our attention to it. For example, he has showed that cognitive and non-cognitive inequalities are present before entering school \citep{heckman2006}. To explain these, and other findings, \citet{cunha2006} suggest that the cognitive level of a child in time \(t\) is a direct function of the experiences of time \(t-1\). Although this sounds logical, it's importance is not understood entirely. The model implies that investments on children are not perfect substitutes at different time points but rather complements. There are specific periods of skill formation where it is much more cost-effective to invest in, which are the early stages of the life course. In fact, when tested against the data, \citet{cunha2006} show that their explanation fits the results of most of the empirical literature very well.

For this reason, children who receive less (or worse) investments have a cumulative disadvantage against children who do receive quality investments. This leads, as early as preschool, to an unequal start in life. That is why the whole debate about narrowing inequality between high and low SES families should start in early education -- it is cheaper than investing at later stages and it brings about the largest returns \citep{elango15}. Despite all of these findings and strategies to reduce the gaps, we still find that the relationship between children's parental education and future destination is present in virtually all empirical studies of social mobility and inequalities \citep{breen1997, breen2007, waldfogel2006, bradbury2015, chetty2016}.

For this reason, it is incomplete to study educational inequalities without considering the prevalence of intergenerational transmission of human and social capital. Social mobility has deepen our knowledge on the relationship between intergenerational transmission of SES and inequality of opportunity. The field of economics of education has also helped to understand that early high-quality education is by far the best investment in terms of cognitive development and acquiring personality skills. Research has constantly found that early education helps low-income children boost their performance to levels similar to their well-off peers \citep{heckman_oecd}. Yet the positive results are not only limited to the cognitive realm: studies have found that that early stimuli is associated with less propensity to criminal activity, more stable relationships, higher levels of persistance, higher income and improved labor market opportunities in late stages in life \citep{heckman_oecd}.

This is not only for the child and his/her family, but to society as a whole as it boosts economic activity, it helps the labor market improve job conditions and maintain a rapid economic growth \citep{hanushek2007}. Since \citet{coleman1966} we've spent most of our time studying the mechanisms through which this inequality comes about. Naturally, we want to do that in order to reverse it and help every child reach its fullest potential. Despite our good intentions, we've concentrated little on the magnitude of the gaps, specially in terms of cognitive abilities. We have very little comparative evidence on which countries have big or small cognitive gaps resulting from SES origins. And even more importantly, we haven't assessed whether policy efforts to reduce inequality have actually had an impact in reducing the achievement gap over time.

@@

From Denmark, for example, the literature has learned a great deal about how to improve social mobility. But that was possible by first learning that Denmark is by far one of the most mobile countries compared to other European countries; this is the case specially for families of low-ses origin \citep{bjorklund2009, jaeger2007}. The important finding came when research emerged the main reason behind their social escalator: it's educational system. For example, research by \citet{esping_waldfogel2012} and \citet{bauchmuller2014}, show that the Danish early education system has important and longstanding impact on improving opportunities. The education system is completely subsidized for all children, otherwise giving opportunities to families who would not have been able to pay. In addition to this, Denmark is recognized as a world-leader in terms of public support for its child-care system as it spends around 2\% of GDP, and has among the highest enrollment rates for children under 6 years old \citep{esping_waldfogel2012}. Moreover, it separates students into different tracks very late compared to other European countries (age 16), something which has been linked to less educational inequality \citep{hanushek_woesmann_tracking}. These two traits make the Danish system effective in promoting equality relative to the average European country. The early schooling experience attempts to level children to the same degree and this process is not stopped by different curricular tracks because it starts at around age 16 when cognitive abilities are less malleable \citep{heckman_oecd}. In short, the importance of this finding, is that we should step back and first study whether the effect is there, how big is it, and then proceed to find the causes behind it.

@@

@@

\subsection{Research questions and hypothesis}

The overaching aim of this paper is to study the evolution of the high/low SES achievement gap in the past 15 years for all PISA participating countries and search for a likely explanation for the evolution.

We develop the subquestions and their corresponding hypothesis separately in more detail.

\begin{enumerate}

\item The seminal work of \citet{reardon2011} suggests that achievement gaps change, and they do so much quicker than we though. After recording a SES gap increase of about 40\% in 30 years, \citet{reardon_portilla} stress that they also found a significant decrease in only 15 years of data, showing how important it is to study the changes in the achievement gap. First, we will concentrate on the evolution of the gap only for 15 year olds (which as we've seen before, there are reasons to think that specific age-groups have seen changes in the achievement gap), which will serve as a comparison to the single year-country snapshot of \citet{anna2016} and the evolution of the kindergarten gap in \citet{reardon_portilla}. Secondly, we will compare the percentage change at which the gap widened/narrowed from the first to the last year available. This will give us a general idea of the overall change over time, and will allow us to compare our estimates to the actual literature. \footnote{Although no study has performed this age-specific achievement gap for comparable tests over such a long time. Our results will serve as comparison for other studies that use age-specific groups, such as 4th graders using PIRLS.}. We posit no specific hypothesis for this question given that it's purely explanatory.

\begin{enumerate}
\item Research question:
\begin{enumerate}
\item Which countries are experiencing changes in the achievement gap?
\end{enumerate}
\end{enumerate}
\end{enumerate}

b)

The widening/narrowing of the achievement gap has a source, which has been often studied to be related to everything from educational spending, income inequality, time allocation to students and preschool enrollment. The literature has concentrated very narrowly on whether the gap is increasing because the top performers are getting ahead, because both are distancing or because the bottom is falling behind. We shall pay particular attention to identifying the rate at which the top/bottom groups are evolving over time. Following that strategy will put us closer to understanding the source of the evolution of the achievement gap.

Research question:
* Is the gap originating because the top is gaining ground, the top is falling behind or because of a dyanmic interaction between the two?

Hypothesis 1:
* The theorical arugment in favor of tracking posits that in countries where there is a high degree of tracking we should expect the top and bottom to be evolving at a similar rate given that tracking is thought to maximize the learning experience of both groups. We expect to find the opposite, i.e for countries with a highly tracked curriculum, we should find more inequality and children from low-ses groups at a greater disadvantage. Conversely, in countries with low tracking we should expect for the achievement gap to be narrowing.

c)

Our study itself is composed of several cross-sectional surveys, which is limited to single cross-sectional snapshots, but given we have 6 times points for most countries we can model the evolution of the achievement gap over time, something not done before in an international context. We want to test whether several dimensions of the tracking setup and vocational enrollment explain the changes in the gaps, something already discussed in \citet{werfhorst_mijs} and \citet{hanushek_woesmann_tracking}.

Research question:
* Are tracking and vocational enrollment related to the evolution of the gap?

Hypothesis 2:
* Building on the previous hypotheiss, tracking should play an important role on the evolution of the achievement gap. We hypothesize that the degree of tracking of an educational system is tightly related to changes in the gap, and the more tracking, the more inequality. Moreover, the more vocational tracking, the less inequality considering that it gives short term returns in terms of labor market opportunities.

d)

We want to investigate whether better performing countries have lower levels of inequality than other countries. \citet{werfhorst_mijs} emphasize that there is empirical evidence that suggests this. This pattern is not so obvious, however. For example, countries with high levels of tracking could maximize student performance, specially the high SES students, raising their overall performance and thus raising the national performance score. But if the bottom performers are not gaining at the same rate, then the achievement gap will inevitable grow resulting in a high performing countries with a widening achievement gap.

Research question:
* Are achievement and inequality correlated? and if so, has the relationship weakeaned/strengthend over time?

Hypothesis 3:
* Building on \citet{werfhorst_mijs}, then there should be a negative relationship between the average country performance and their level of inequality. Countries with high performance should have less inequality and countries with low performance should have greater inequality
